---
title: "Interactive Data Visualization in `plotly`"
subtitle: ""
author: "Berry Consultants, LLC"
date: ""
link-citations: yes
linkcolor: blue
output:
  bookdown::html_document2:
    fig_caption: yes
    theme: flatly # sandstone # spacelab # flatly
    highlight: pygments
    toc: TRUE
    toc_depth: 4
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
---

```{r initial_setup, include = FALSE, echo = FALSE, results = "hide", message = FALSE, warning = FALSE, error = FALSE}
knitr::opts_chunk$set(error = FALSE, comment = NA, warning = FALSE, message = FALSE)
options(knitr.kable.NA = '', knitr.table.format = "html", scipen = 999)
```


```{r, echo = FALSE, results = "hide"}
source('./utils_plots.R')

weeks_to_months = 6/26
cols = brewer.pal(9, 'Set1')


data = read_csv(file = 'data_ordinal.csv') %>% 
  mutate(TRTPN = factor(TRTPN))
key_trt = levels(data$TRTPN)

digits = 0
n_digits = if_else(digits == 0, "%s", paste0("%0.", digits, "f"))
K = data %>% 
  filter(!is.na(RESP180)) %>% 
  select(RESP180) %>% 
  n_distinct


# Priors for the longitudinal model
long_prior = list(beta = rep(0, K), # priors for p30 
                  alpha = matrix(rep(0, K*K), 
                                 K, K, byrow = T))

imp_pars = get_imputation_pars(data, long_prior, key_trt)


# Accrual profile
accr_profile = NULL
accr_profile$peak_rate = 12 * weeks_to_months # in weeks
accr_profile$ramp_up = 24 / weeks_to_months # in weeks


# STAN data
stan_data <- list(N = nrow(data %>% filter(OTC180FL == 1)),
                  D = K, 
                  X = as.numeric(data %>% filter(OTC180FL == 1) %>% pull(TRTPN)) - 1, 
                  y = data %>% filter(OTC180FL == 1) %>% pull(RESP180) + 1
)


# Run STAN model
model_seed = 123
n_iters = 5000
burnin = 2500
n_thin = 2
n_chains = 4
samp_size = n_chains * (n_iters - burnin) / n_thin
set.seed(model_seed)
fit <- stan(file = '../Ordinal Regression/ord_logistic.stan',
            data = stan_data,
            chains = n_chains,
            warmup = burnin,
            iter = n_iters,
            thin = n_thin,
            seed = model_seed) %>%
  extract()


# Extract model output
samples = fit$theta %>%
  cbind(fit$c) %>%
  colnames_inplace(c('theta', sprintf('c%s', 1:(K - 1))))


# Show the posterior distribution of the class probabilities under the control group
p_ctr_true = c(0.07, 0.2, 0.28, 0.2, 0.15, 0.1)
p_ctr_post = cbind(rep(0, samp_size), inv_logit(samples[,2:K]), rep(1, samp_size)) %>% 
  apply(1, diff) %>% 
  t() %>% 
  colnames_inplace(sprintf('p%s', 0:(K - 1)))
p_ctr_est = p_ctr_post %>% 
  as_tibble %>% 
  pivot_longer(cols = everything(), names_to = 'param') %>% 
  group_by(param) %>% 
  summarize(mean = mean(value), 
            low95 = quantile(value, 0.025), 
            upp95 = quantile(value, 0.975), 
            low50 = quantile(value, 0.25), 
            upp50 = quantile(value, 0.75)) %>% 
  mutate(true = p_ctr_true, 
         param = as.numeric(str_remove(param, 'p')))
```


# General Features


Plotly is a scientific graphing library that can be used for Python, R, MATLAB, as well as other programming languages. 
We focus here on plotly for `R`, whose documentation can be found [here](https://plotly.com/r/).


Luckily, `plotly` works very similarly to `ggplot2`. 
Each element of the plot has to be specified in a separate *layer* (or, in the case of `plotly`, in a separate *trace*).
A good function to start experimenting is `ggplotly()`. 
This function, when applied to a `ggplot` object, tries to render the graph as an interactive plotly-like visualization. 
However, `ggplotly()` can be hard to customize to more specific needs.


In this walkthrough, interactive visualization examples will be provided.
Hopefully these are a starting point for people who are interested in developing interactive visualizations.  


# Heatmaps


```{r, echo = FALSE}
plt_heat = imp_pars$p90_ctr %>%
  as_tibble() %>% 
  colnames_inplace(as.character(0:(K - 1))) %>% 
  mutate(RESP30 = as.character(0:(K - 1)), .before = 1) %>% 
  pivot_longer(
    cols = -RESP30,
    names_to = "RESP90",
    values_to = "transitions"
  )
```


Consider an ordinal endpoint measured over time and the following transitions from state at 30 days to state at 90 days. 


```{r, results = 'hide'}
head(plt_heat, 6)
```
  
  
```{r, echo = FALSE}
plt_heat %>% 
  head(n = 6) %>% 
  knitr::kable(booktabs = TRUE) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```


We can show the transition probability matrix with a heatmap whose cells are colored according to the normalized probabilities for each fixed row.
Thus, the colors can be interpreted as the probability of going to each state at day 90, conditional on being at a certain state at day 30.


Note the option `hovertemplate` that is highly flexible and allows to customize the text that appears on hover.


```{r}
plt_heat %>% 
  group_by(RESP30) %>% # we calculate the normalized proportions by row to color the cells
  mutate(p_transitions = transitions / sum(transitions), 
         p_transitions = if_else(is.na(p_transitions), 0, p_transitions)) %>% 
  ungroup() %>% 
  mutate(label = sprintf(n_digits, transitions)) %>%
  plot_ly(
    x = ~RESP90, 
    y = ~RESP30,
    z = ~p_transitions, 
    text = ~label, 
    type = "heatmap", 
    hovertemplate = paste0('State <b>%{y}</b> -> State <b>%{x}</b>',
                           '<br><b>%{text}</b><extra></extra>'),
    colors = "Greens", 
    showscale = FALSE
  ) %>% 
  add_annotations(showarrow = FALSE, 
                  font = list(size = 16)) %>%
  layout(title = '', 
         xaxis = list(title = 'Response at 90 days', 
                      showgrid = FALSE, 
                      zeroline = FALSE), 
         yaxis = list(title = 'Response at 30 days', 
                      autorange = "reversed", 
                      # scaleanchor = "x", 
                      # scaleratio = 1, 
                      showgrid = FALSE, 
                      zeroline = FALSE)
  )
```


# Line Plots


```{r, echo = FALSE}
plt_accr = data %>% 
  mutate(TIMESINCERAND = as.numeric(difftime(RANDDT, min(RANDDT), units = 'weeks'))) %>% 
  arrange(TIMESINCERAND) %>% 
  mutate(Subject = row_number(), 
         EXPACCR = Lambda(t = TIMESINCERAND, peak_rate = accr_profile$peak_rate, 
                          ramp_up = accr_profile$ramp_up)) %>% 
  select(RANDDT, Subject, EXPACCR)
```


Let's say that we have the following variables indicating randomization date, subject number, and expected accrual at that date. 


```{r, results = 'hide'}
head(plt_accr, 6)
```
  

```{r, echo = FALSE}
plt_accr %>% 
  head(n = 6) %>% 
  knitr::kable(booktabs = TRUE) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```


We can visualize the accrual with a simple plot containing two lines.
One of the two lines is the cumulative number of subjects and therefore it has the `shape = 'hv'` option to make it look like a step function.


```{r}
plt_accr %>% 
  plot_ly(
    type = 'scatter', 
    mode = 'lines', 
    line = list(color = cols), 
    hoverinfo = 'x+y') %>% 
  add_trace(x = ~RANDDT, 
            y = ~Subject, 
            name = 'observed', 
            line = list(shape = 'hv')) %>% 
  add_trace(x = ~RANDDT, 
            y = ~EXPACCR, 
            name = sprintf('%.2f per week', accr_profile$peak_rate)) %>%
  layout(title = 'Observed vs simulated accrual',
         xaxis = list(title = 'Randomization date'),
         yaxis = list(title = 'Cumulative number of subjects'), 
         legend = list(orientation = "h",   # show entries horizontally
                       xanchor = "center",  # use center of legend as anchor
                       x = 0.5, 
                       y = 1)
  )
```


Given a point estimate (i.e., posterior mean) with its associated $95\%$ and $50\%$ credible intervals, we can show it with a shaded region.


```{r, results = 'hide'}
p_ctr_est
```
  
  
```{r, echo = FALSE}
p_ctr_est %>% 
  knitr::kable(booktabs = TRUE) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r}
p_ctr_est %>% 
  # this trick makes the right part of the plota flat line
  rbind(tail(p_ctr_est, 1) %>% mutate(param = param + 1)) %>% 
  plot_ly(x = ~param, y = ~upp95, 
          type = 'scatter', mode = 'lines',
          line = list(color = 'transparent', shape = 'hv'),
          showlegend = FALSE, name = '95% CI') %>% 
  add_trace(y = ~low95, 
            fill = 'tonexty', fillcolor='rgba(0,100,80,0.3)', 
            name = '95% CI') %>% 
  add_trace(y = ~upp50, 
            name = '50% CI') %>% 
  add_trace(y = ~low50, 
            fill = 'tonexty', fillcolor='rgba(0,100,80,0.4)', 
            name = '50% CI') %>% 
  add_trace(y = ~mean, 
            line = list(color='rgb(0,100,80)'),
            name = 'Mean') %>% 
  add_trace(y = ~true, 
            name = 'True', 
            line = list(color='black')) %>% 
  layout(title = "",
         xaxis = list(title = "", 
                      range = c(0, K),
                      ticktext = list("p0", "p1", "p2", "p3", "p4", "p5"), 
                      tickvals = list(0.5, 1.5, 2.5, 3.5, 4.5, 5.5)),
         yaxis = list(title = "", 
                      range = c(0, 0.5)))
```



# Bar Charts


Consider again data collected on an ordinal endpoint that ranges from $0$ to $`r K-1`$.
```{r, results = 'hide'}
head(data, 6)
```
  
  
```{r, echo = FALSE}
data %>% 
  head(n = 6) %>% 
  knitr::kable(booktabs = TRUE) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```


A first possible to visualize the data is with the stacked barplot below.

```{r}
data %>%
  filter(OTC180FL == 1) %>% 
  group_by(TRTPN, RESP180) %>% 
  summarise(n = n()) %>% 
  mutate(freq = n / sum(n)) %>% 
  right_join(list(
    TRTPN = key_trt,
    RESP180 = 0:(K-1)) %>%
      cross_df(), 
    by = c('TRTPN', 'RESP180')) %>%
  mutate(n = ifelse(is.na(n), 0, n), 
         freq = ifelse(is.na(freq), 0, freq)) %>% 
  arrange(TRTPN, RESP180) %>% 
  group_by(TRTPN) %>% 
  mutate(freq_sum = cumsum(freq)) %>% 
  plot_ly(x = ~freq, 
          y = ~TRTPN, 
          color = ~factor(RESP180, levels = 1:K-1),
          text = ~as.character(RESP180),
          colors = brewer.pal(K, 'RdBu')[K:1],
          type = 'bar', 
          hovertemplate = paste0('<b>%{y}</b><br>', 
                                 'Proportion with response = %{text}:<br>',
                                 '%{x:.2f}<extra></extra>')
  ) %>% 
  layout(title = 'Frequency of the response at 180 days',
         xaxis = list(title = 'Proportion of subjects'),
         yaxis = list(title = ''), 
         legend = list(title = list(text='<b>Response</b>'), 
                       traceorder = 'normal', 
                       orientation = "h",  # show entries horizontally
                       xanchor = "center", # use center of legend as anchor
                       x = 0.5, 
                       y = 1.02), 
         barmode = 'stack'
  )
```


A similar representation can be obtained via a cumulative distribution function superimposed on a barplot. 


```{r}
data %>%
  filter(OTC180FL == 1) %>% 
  group_by(TRTPN, RESP180) %>% 
  summarise(n = n()) %>% 
  mutate(freq = n / sum(n)) %>% 
  right_join(list(
    TRTPN = key_trt,
    RESP180 = 0:(K-1)) %>%
      cross_df(), 
    by = c('TRTPN','RESP180')) %>%
  mutate(TRTPN = factor(TRTPN), 
         n = ifelse(is.na(n), 0, n), 
         freq = ifelse(is.na(freq), 0, freq)) %>% 
  arrange(TRTPN, RESP180) %>% 
  group_by(TRTPN) %>% 
  mutate(freq_sum = cumsum(freq), 
         imp_month_plot = ifelse(TRTPN == key_trt[1], RESP180, RESP180)) %>% 
  plot_ly(x = ~RESP180,
          y = ~freq, 
          color = ~TRTPN,
          text = ~as.character(TRTPN),
          colors = 'Dark2',
          type = 'bar',
          hovertemplate = paste0('<b>%{text}</b><br>',
                                 'Proportion with response = %{x}:<br>',
                                 '%{y:.2f}<extra></extra>')
  ) %>% 
  add_trace(x = ~imp_month_plot, 
            y = ~freq_sum, 
            color = ~TRTPN,
            text = ~as.character(TRTPN),
            type = 'scatter', 
            mode = 'lines', 
            line = list(shape = 'hv'), 
            showlegend = F, 
            hovertemplate = paste0('<b>%{text}</b><br>',
                                   'Cumulative proportion with response <= %{x}:<br>',
                                   '%{y:.2f}<extra></extra>')) %>% 
  layout(title = 'Frequency of the response at 180 days',
         xaxis = list(title = 'Proportion of subjects'),
         yaxis = list(title = ''), 
         legend = list(orientation = "h",  # show entries horizontally
                       xanchor = "center", # use center of legend as anchor
                       x = 0.5, 
                       y = 1), 
         barmode = 'group', 
         bargap = 0.3, 
         bargroupgap = 0.05
  )
```


# Histograms


Say that we have posterior samples from several parameters. 


```{r, results = 'hide'}
head(samples, 6)
```
  
  
```{r, echo = FALSE}
samples %>% 
  head(n = 6) %>% 
  knitr::kable(booktabs = TRUE) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```


We can visualize the distribution of them with histograms whose colors represent the different parameters.

```{r}
samples %>% 
  as_tibble %>% 
  select(contains('c')) %>% 
  pivot_longer(cols = everything(), names_to = 'cutpoint') %>% 
  plot_ly(x = ~value, color = ~factor(cutpoint), 
          alpha = 0.6, type = "histogram", 
          histnorm = "probability") %>% 
  layout(barmode = "overlay", 
         xaxis = list(title = 'Cutpoints'), 
         yaxis = list(title = 'density'), 
         legend = list(orientation = "h",  # show entries horizontally
                       xanchor = "center", # use center of legend as anchor
                       x = 0.5, 
                       y = 1))
```




<!-- ```{r} -->
<!-- fig <- plot_ly( -->
<!--     type = "sankey", -->
<!--     orientation = "h", -->

<!--     node = list( -->
<!--       label = c("A1", "A2", "B1", "B2", "C1", "C2"), -->
<!--       color = c("blue", "blue", "blue", "blue", "blue", "blue"), -->
<!--       pad = 15, -->
<!--       thickness = 20, -->
<!--       line = list( -->
<!--         color = "black", -->
<!--         width = 0.5 -->
<!--       ) -->
<!--     ), -->

<!--     link = list( -->
<!--       source = c(0,1,0,2,3,3), -->
<!--       target = c(2,3,3,4,4,5), -->
<!--       value =  c(8,4,2,8,4,2) -->
<!--     ) -->
<!--   ) -->
<!-- fig <- fig %>% layout( -->
<!--     title = "Basic Sankey Diagram", -->
<!--     font = list( -->
<!--       size = 10 -->
<!--     ) -->
<!-- ) -->

<!-- fig -->
<!-- ``` -->


